---
title: "Tensor diagrams"
description: |
  A new article created using the Distill format.
author:
  - name: Piotr Migdał
    url: https://p.migdal.pl
    affiliation: Quantum Flytrap
    affiliation_url: https://quantumgame.io
date: "`r Sys.Date()`"
bibliography: references.bib
output: distill::distill_article
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

Idea:

- Linear algebra.
- Copy operation.
- Stochastic models.
- Tensors in the sense of TensorFlow and torch.tensor.


Tensor diagrams where invented by Penrose[@penrose_applications_1971].
For the first contact with tensor diagrams I suggest glaring at the beautiful diagrams[@bradley_matrices_2019]. If you have some background in quantum mechanics, go for a short introduciton[@biamonte_tensor_2017] or a slightly longer [@bridgeman_hand-waving_2017].

For a complete introduction I suggest book [@coecke_picturing_2017] or lecture notes [@biamonte_lectures_2020].

Tensor diagrams are popular quantum state decomposition for condensed matter physics[@orus_practical_2014,@verstraete_matrix_2008].

Using a script from [Tensor diagram notation in D3.js - JSFiddle](https://jsfiddle.net/stared/8huz5gy7/).



A vector:

```{r fig.height=2}
library(r2d3)
data <- jsonlite::read_json("diagrams/example.json")
r2d3(data, script = "tensorDiagram.js", css = "tensorDiagram.css")
```




## You may have seen tensor diagrams

Feynman Diagrams[@kaiser_physics_2005] (for unbounded operators on infitely-dimensional Hilbert spaces) and quantum computing.

TODO: Do LSTM diagrams qualify?

## Hidden Markov Models

Hidden Markov Models[@rabiner_tutorial_1989].

## More 


## Footnote

Written in [R Markdown Distill](https://rstudio.github.io/distill).